GEE models

### 2016-04-15 doc/manuscript.Rmd
Many changes to manuscript (extra analysis, figs, dysgly, edits) before sending to Tony

---
bibliography: 
    - /home/luke/Documents/bibliography/bib/bib-db/nefaDiabetes.bib
    - /home/luke/Documents/bibliography/bib/bib-db/myrefs.bib
csl:
    - /home/luke/Documents/bibliography/styles/vancouver.csl
output:
    md_document:
        variant: markdown
    word_document:
        reference_docx: diabetes-care-template.docx
        fig_height: 6 
        fig_width: 5
        fig_caption: true
---

@@

Word count: / 4000

Table and figure count: / 4

@@

# Abstract

@@

TODO:
- Bootstrap for the longitudinal table data... like instead of an ANOVA/GEE for
the p-value...?
- What about adiponectin?
- Precise followup time for each participant?
- Interaction by sex or ethnicity

@@

```{r setup, message=FALSE}
source('.Rprofile')
suppressMessages(run_setup())
ds <- load_data()
```

@@

tab_basic <- tab('basicChar', 'Basic characteristics of the PROMISE participants at each of the 3 clinic visits.')
fig_dist <- fig('nefaDist', 'Concentrations (nmol/mL) of each non-esterified fatty acid in PROMISE participants at the baseline visit (2004-2006)')
fig_gee <- fig('gee', 'Longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function using generalized estimating equations over the 6 years in the PROMISE cohort.  Adjusted for {{covariates}}. Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid.  P-values presented were adjusted for the false discovery rate.')

sfig_consort <- sfig('consort', 'CONSORT diagram of sample size at each examination visit.')
stab_qic <- stab('qic', 'Comparing generalized estimating equation models adjusting for different covariates using Quasi-Likelihood Information Criterion.')
sfig_corr <- sfig('corr', 'Pearson correlation heatmap of non-esterified fatty acids (nmol/mL) and basic PROMISE participant characteristics for the baseline visit (2004-2006)')
sfig_gee <- sfig('gee', 'Unadjusted generalized estimating equation modeling of the longitudinal association of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function over 6 years in the PROMISE cohort. Only adjusted for time. Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid. P-values presented were adjusted for the false discovery rate.')
stab_gee <- stab('gee', 'Longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function using generalized estimating equations over the 6 years in the PROMISE cohort.  Adjusted for {{covariates}}. Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid.  P-values presented were adjusted for the BH false discovery rate.')
stab_gee_unadj <- stab('gee_unadj', 'Unadjusted generalized estimating equations models of the longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function over the 6 years in the PROMISE cohort.  Adjusted for time only. Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid.  P-values presented were adjusted for the BH false discovery rate.')
sfig_lcmm <- sfig('lcmm', 'Latent class mixed model analysis to identify individual classes of trajectories for log(ISSI-2) over the 6 years in the PROMISE cohort.')
sfig_plsda_loadings <- sfig('plsda_loadings', 'Pattern loadings from partial least squares discriminant analysis to identify potential clusters of NEFA composition within the classes extracted from the latent class mixed model in 463 participants from the baseline PROMISE visit (2004-2006). The percent explained variance of each component is shown in brackets on each axis.')
sfig_plsda_groups <- sfig('plsda_groups', 'Clustering of extracted components from the partial least squares discriminant analysis on the classes extracted from the latent class mixed model (LCMM) in 463 participants from the baseline PROMISE visit (2004-2006). The percent explained variance of each component is shown in brackets on each axis.')
```

@@

Total non-esterified fatty acids (NEFA) have been well documented to influence
the pathogenesis of type 2 diabetes mellitus. Experimental work has shown that
exposure of cells or animals to high concentrations of NEFA can induce insulin
resistance in insulin-sensitive tissue such as the muscle or liver and impair
beta-cell production of insulin, and cause hyperglycemia [@Rachek2014a;@Cnop2008a].
Epidemiological and clinical studies have observed similar findings, showing
that elevated total plasma NEFA associates with risk for diabetes [@Daniele2014a {{more}}].

Most previous experimental work uses palmitic acid (16:0), oleic acid (18:1n-9),
or a standard oil such as soybean {{confirm, cite}} and generalize as total
NEFA. However, fatty acids comprise multiple classes of molecules with diverse 
physiological functions, with few studies comparing different fatty acids. For
instance, one study on beta-cell {{plates? lines?}} suggests that
eicosapentaenoic acid (20:5n-3) can protect against the lipotoxic effect of
palmitic acid {{xiao? cite}}.

Prospective with diabetes [@Ilyasova2010a]

Lifestyle interventions aimed at managing or preventing diabetes show a
concomitant reduction in NEFA and insulin resistance {{cite}}.

Lack of longitudinal data is particularly important given that NEFA release from
adipocytes is inhibited by insulin, so if there is IR, more NEFA are released,
making it difficult to disentangle which caused which.

Despite the fairly extensive literature on NEFA and diabetes, there are
substantial research gaps. Importantly, most of the experimental work using
individual NEFA species has been in animal and cell lines or if in humans has
been acute, infusion studies. For epidemiological studies, few have examined the
role of total, but not individual, NEFA on the pathogenesis of diabetes and
fewer still have been longitudinal, which is key to understanding causal
associations and given the long timescale in diabetes pathophysiology.
Presently, no longitudinal study has been conducted that has examined the role
of the composition of the NEFA fraction on the pathogenesis of diabetes. Our
research objective was to examine the role of the NEFA composition
on insulin sensitivity and beta-cell function using longitudinal and clustering
techniques. Some specific hypotheses included that higher palmitic acid while
lower polyunsaturated fatty acids such as eicosapentaenoic acid would associate
with lower insulin sensitivity and beta-cell function.

@@

Participants from London and Toronto, Canada, were recruited into the 
Prospective Metabolism and Islet cell Evaluation (PROMISE) cohort. Eligibility 
for recruitment into PROMISE included having one or more risk factors for type 2
diabetes mellitus, such as obesity, hypertension, family history of diabetes, 
and/or a history of gestational diabetes or birth of a macrosomic infant. 
Participants aged 30 years and older (n=726) conducted the baseline visit
between 2004-2006. Follow-up examinations occur every three years, with three
examination visits completed to date (2004-2006, 2007-2009, and 2010-2013).  The
current study used data on participants who did not have diabetes at baseline,
who returned for one or more follow-up examinations, and who had samples
available for fatty acid measurements (n=477).  A diagram of the sample size at
each visit is shown in `r cite_sf('consort')`.  At each examination,
participants undergo metabolic characterization, anthropometric measurements,
and questionnaires on lifestyle and sociodemographics.  Research ethics approval
was obtained from Mount Sinai Hospital and the University of Western Ontario. 
Data collection methods were standardized across the 2 centres and research
nurses were centrally trained.

@@

NEFA composition was quantified using stored fasting serum samples from the 
baseline visit, which had been frozen at -70°C for 4-6 years and had not been 
exposed to any freeze-thaw cycles.  Serum fatty acids have been documented to be
stable at these temperatures for up to 10 years [@Matthan2010a]. A known amount 
of heptadecanoic acid (17:0) {{confirm}} was added to the serum as an internal 
standard prior to extracting total lipids according to the method of Folch 
[@Folch1957a].  Each serum lipid fraction was isolated using thin layer 
chromatography, which were visualized under UV light after lightly spraying with
8-anilino-1-naphthalene sulfonic acid (0.1% wt/vol) and then converted to fatty
acid methyl esters with 14% boron trifluoride in methanol at 100°C for 1 h. 
Fatty acid methyl esters were separated and quantified using a Varian-430 gas
chromatograph (Varian, Lake Forest, CA, USA) equipped with a Varian Factor Four
capillary column and a flame ionization detector, which were injected in
splitless mode. Fatty acid concentrations (nmol/ml) were calculated by
proportional comparison of gas chromatography peak areas to that of the internal
standards [@Nishi2014a; @Chen2011a].  There were 22 fatty acids measured in the
NEFA fraction.

@@

For the primary analysis, generalized estimating equation (GEE) models 
[@Zeger1986a] were used to determine the longitudinal associations between the 
outcome variables and the predictor variables.  Given the longitudinal design 
and auto-regressive (AR1) correlation matrix was chosen for the GEE models, 
though other matrices (eg. exchangeable) had similar fit (data not shown). GEE
is well suited to cohort studies given it's capacity to handle missed visits.
The predictor variables and continuous covariates were scaled (mean centered and
standardized). The NEFA variables were classified as *time-independent* (held
constant) as they were measured only at the baseline visit, while the outcome
variables and covariates were set as *time-dependent*. No imputation was
conducted on missing values.

Covariates selection was based on previous literature, directed acyclic graph 
[@Greenland1999a] recommendations, and quasi-likelihood information criterion 
(QIC). `r cite_st('qic')` shows the covariates compared using QIC and `r 
cite_sf('dag')` shows the specified directed acyclic graph. The final GEE model 
we selected differed between insulin sensitivity and beta-cell function models. 
For insulin sensitivity, we adjusted for time, sex, ethnicity, baseline age, WC,
ALT, MET, alcohol intake, and family history of diabetes (M6). For beta-cell
function, the last few models were comparable in performance, so we chose the
simplest model (M8), adjusting for time, sex, ethnicity, baseline age, WC, ALT,
and family history of diabetes. After scaling, log-transforming, and
exponentiating the GEE estimates, the GEE results are interpreted as an expected
percent difference in the outcome variable for every SD increase in the
predictor variable given the covariates are held constant. While models with a
time interaction did not perform better than others for the QIC (M6.1 and M8.1
in `r cite_st('qic')`, we ran models with a time by predictor term for each of
the NEFA to test if there was a time interaction. Given that TG is a risk factor
for diabetes and that NEFA contribute to TG production, TG may act as a mediator
between NEFA and the outcomes. However, only a subset of the participants at the
6-yr visit had clinically measured TG. To determine the role of TG in the
association between NEFA and the outcomes, TG was included in the GEE model as a
sensitivity analysis.

While GEE controls for the longitudinal design of the data, a limitation is that
it can not analyze the inherent multivariate nature of the composition of the 
NEFA fraction. As such, after running the GEE models for the longitudinal 
analysis, partial least squares discriminant analysis (PLS-DA) was used to
identify the patterns of NEFA composition against beta-cell function (using ISSI-2,
which has most of the associations in the GEE models). Since PLS-DA can't handle
longitudinal data, the longitudinal data was maximized by extracting latent
trajectories of beta-cell function over the 6 years using latent class mixed
models (LCMM). A limitation of these methods is that they can not be adjusted
for covariates.

All analyses were performed using R `r getRversion()` [@Rbase], along with the R
packages geepack `r packageVersion('geepack')` for GEE [@Hoejsgaard2006a], caret
`r packageVersion('caret')` for PLS-DA, and lcmm `r packageVersion('lcmm')` for
LCMM. The R code for this manuscript is available at {{link}}. Results were
statistically significance at p<0.05, after adjusting for multiple testing using
the False Discovery Rate [@Benjamini1995a].

@@

```{r variableSet}
gee_df <- get_gee_data(ds)

@@

```{r lcmmPls}
# include also the 3-yr and 6-yr data for training/testing.
lcmm_results <- analyze_lcmm(ds)
lcmm_data <- combine_lcmm_data(ds, lcmm_results)
plsda_results <- analyze_plsda(lcmm_data)
```

```{r inlineResults}
percent.nefa <- calculate_percent_nefa_contribution(ds) 
misclass.n.percent <- calculate_plsda_misclass(plsda_results)
percent.change.outcomes <- calculate_outcomes_pct_change(ds)
lcmm.groups <- calculate_ngroups_lcmm(lcmm_results)
```

@@

The primary outcome variables had a median decline between 
`r percent.change.outcomes` over the 6 years in this subset of the PROMISE cohort.

`r cite_f('nefaDist')` shows the distribution of the NEFA composition. Four 
individual NEFA made up the vast majority (`r sum(pct_nefa$pct)`%) of the total
NEFA fraction. The largest contributors were `r pctne$c[1]`, `r pctne$c[2]`, 
`r pctne$c[3]`, and `r pctne$c[4]`. All individual NEFA as well as the total
fraction had weak to no correlations (r<0.3) with the basic characteristics (see
`r cite_sf('corr')`).

@@

Several associations were seen in the unadjusted GEE models 
(`r cite_sf('gee')`), particularly for the concentration data. Full model
adjustment attenuated most of these associations, where total NEFA, 16:0, and
18:2n-6 (all as nmol/mL) had negative associations between IGI/IR and ISSI-2
(see `r cite_f('gee')`). The magnitude of association for each of these
variables was fairly consistent for each beta-cell function measure. For every
one SD decrease in any of these three NEFA variables, there was a predicted
{{magnitude}}% lower IGI/IR and a {{magnitude}}% lower ISSI-2 at each clinic
visit. None of the NEFA variables as a mol% had any association with the
outcomes. Adjusting for TG attenuates all associations with IGI/IR and ISSI-2
(data not shown), though sample size at the 6 year visit is restricted to n=128
in those who had TG measured. There were no significant interaction effects
between time and the individual NEFA on any of the outcome measures (data not
shown). Raw values from the GEE models are shown in `r cite_st('gee_unadj')` for
unadjusted models and `r cite_st('gee')` for fully adjusted models.

@@

Three latent classes were extracted from LCMM, with `r lcmm_grps[2,2]` 
participants in the high, `r lcmm_grps[1,2]` in the middle, and `lcmm_grps[3,2]`
in the low beta-cell function groups. The trajectories of the three groups are
shown in `r cite_sf('lcmm')`. These groups were used as the response variables
in the PLS-DA analysis. The PLS-DA results show a very poor discriminatory
abililty of the NEFA composition in classifying participants into the correct
beta-cell function latent class. Only `r misclass_npct[2,1]` 
(`r misclass_npct[2,4]`) participants were correctly classified, primarily into the
middle group. The specific NEFA composition was not able to accurately classify
participants who had low or high beta-cell function.
`r cite_sf('plsda_loadings')` shows the loadings of individual NEFA by the
extracted component and `r cite_sf('plsda_groups')` shows the clustering of the
components by ISSI-2 latent class.

@@

In a Canadian population of adults who are at-risk for diabetes, we found that
higher total NEFA independently predicted a lower beta-cell function over 6
years. In particular, the specific composition of the NEFA fraction had no
association with any of the metabolic outcomes within this cohort. These
findings suggest that the size of the total fraction, rather than the
composition, influences the pathogenesis of diabetes.
Given we found associations with the nmol/mL data, but none in the mol% data,
this suggests to us that the total size of the NEFA fraction is more important
in the pathogenesis of diabetes than is the specific FA composition. A
limitation of the mol% data is the inherent dependency between the FA as an
increase in one FA by definition means other FA must decrease. In the context of
this dependency, the associations we found for 16:0 and 18:2n-6 as nmol/mL
suggests that these NEFA may be contributing to the harmful association seen for
the total NEFA fraction, irrespective of changes in the other NEFA.

While future studies are needed to confirm these findings, our results suggest
that measurement of the NEFA composition may not be warranted and that the
simple, easy to measure total NEFA may be the measure most appropriate for
further analysis. {{reword}}

>>>>

---
bibliography: 
    - /home/luke/Documents/bibliography/bib/bib-db/nefaDiabetes.bib
    - /home/luke/Documents/bibliography/bib/bib-db/myrefs.bib
csl: /home/luke/Documents/bibliography/styles/dependent/diabetes-care.csl
output:
    word_document:
        reference_docx: diabetes-care-template.docx
        fig_height: 6 
        fig_width: 5
        fig_caption: true
        keep_md: true
---

@@

Word count: 3527 / 4000

Table and figure count: 3 / 4

@@

@@

TODO:
- Bootstrap for the longitudinal table data... like instead of an ANOVA/GEE for
the p-value...?
- What about adiponectin?
- Precise followup time for each participant?
- [X] Interaction by sex or ethnicity
- [X] Interaction by time
- [X] Sensitivity analysis with TG
- [X] PLS-DA with dysglycemia (for anyone who got it over the years).

@@

```{r setup, message=FALSE, echo=FALSE}
source('.Rprofile')
suppressMessages(run_setup())
ds <- load_data()
ds0yr <- filter(ds, VN == 0)
```

@@

tab_basic <- tab('basicChar', 'Basic characteristics of the PROMISE participants at each of the 3 clinic visits.')
fig_dist <- fig('nefaDist', 'Concentrations (nmol/mL) of each non-esterified fatty acid in PROMISE participants at the baseline visit (2004-2006).')
fig_gee <- fig('gee', 'Longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function using generalized estimating equations over the 6 years in the PROMISE cohort.  Adjusted for time, sex, ethnicity, baseline age, WC,
ALT, and family history of diabetes (plus physical activity and alcohol intake for beta-cell function models). Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid. P-values were adjusted for the BH false discovery rate, presented as the dot size.')

sfig_consort <- sfig('consort', 'CONSORT diagram of sample size at each examination visit.')
stab_qic <- stab('qic', 'Comparing generalized estimating equation models adjusting for different covariates using Quasi-Likelihood Information Criterion.')
sfig_corr <- sfig('corr', 'Pearson correlation heatmap of non-esterified fatty acids (nmol/mL) and basic PROMISE participant characteristics for the baseline visit (2004-2006). Darkness of the colour indicates the size of the correlation, with blue indicating positive and orange indicating negative correlations.')
sfig_gee <- sfig('gee', 'Unadjusted generalized estimating equation modeling of the longitudinal association of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function over 6 years in the PROMISE cohort. Only adjusted for time. Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid. P-values were adjusted for the BH false discovery rate, presented as the dot size.')
stab_gee <- stab('gee', 'Longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function using generalized estimating equations over the 6 years in the PROMISE cohort.  Adjusted for time, sex, ethnicity, baseline age, WC,
ALT, and family history of diabetes (plus physical activity and alcohol intake for beta-cell function models). Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid.  P-values were adjusted for the BH false discovery rate, indicated by asterisk.')
stab_gee_unadj <- stab('gee_unadj', 'Unadjusted generalized estimating equations models of the longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function over the 6 years in the PROMISE cohort.  Adjusted for time only. Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid.  P-values were adjusted for the BH false discovery rate, indicated by asterisk.')
sfig_lcmm <- sfig('lcmm', 'Latent class mixed model analysis to identify individual classes of trajectories for log(ISSI-2) over the 6 years in the PROMISE cohort.')
sfig_plsda_groups <- sfig('plsda_groups', 'Clustering of extracted components from the partial least squares discriminant analysis on the classes extracted from the latent class mixed model (LCMM) in 463 participants from the baseline PROMISE visit (2004-2006). The percent explained variance of each component is shown in brackets on each axis. {{colour}}, {{colour}}, {{colour}} lines indicate participants classified as high, middle, and low for beta-cell function, respectively, from the LCMM analysis.')
sfig_plsda_loadings <- sfig('plsda_loadings', 'Pattern loadings from partial least squares discriminant analysis to identify potential clusters of NEFA composition within the classes extracted from the latent class mixed model in 463 participants from the baseline PROMISE visit (2004-2006). The percent explained variance of each component is shown in brackets on each axis.')
sfig_plsda_groups_dysgly <- sfig('plsda_groups_dysgly', 'Clustering of extracted components from the partial least squares discriminant analysis for dysglycemia (IFG, IGT, DM) conversion status over the 6-years in the participants from the baseline PROMISE visit (2004-2006). The percent explained variance of each component is shown in brackets on each axis. {{colour}} lines indicate dysglycemia conversion or maintanence and {{colour}} lines indicate no dysglycemia status.')
sfig_plsda_loadings_dysgly <- sfig('plsda_loadings_dysgly', 'Pattern loadings from partial least squares discriminant analysis to identify potential clusters of NEFA composition for dysglycemia (IFG, IGT, DM) conversion status over the 6-years in the participants from the baseline PROMISE visit (2004-2006). The percent explained variance of each component is shown in brackets on each axis.')
```

@@

Total non-esterified fatty acids (NEFA) have been well documented to influence 
the pathogenesis of type 2 diabetes mellitus. Experimental work has shown that 
exposure to high concentrations of NEFA can induce insulin resistance in
insulin-sensitive tissue such as the muscle or liver and impair beta-cell
production of insulin [@Rachek2014a;@Cnop2008a]. Epidemiological and clinical
studies have observed similar findings, showing that elevated total plasma NEFA
associates with risk for diabetes [@Daniele2014a;@Salgin2012a].

Much of the previous experimental work frequently use fatty acids such as
palmitic acid (16:0) or oleic acid (18:1n-9), or use a standard oil such as soybean
oil as the exposure and generalize as total NEFA. However, fatty acids comprise
multiple classes of molecules with diverse physiological functions, with few
studies comparing different fatty acids. For instance, one study suggests that
eicosapentaenoic acid (20:5n-3) can protect against the lipotoxic effect of
palmitic acid in the beta-cells [@Xiao2006].

Despite a sizable literature studying the role of NEFA in diabetes, there are 
important gaps in the research. In particular, there have been few long term 
studies in humans that have examined a range of NEFA species. Most studies have 
been on animals and cells, have been short term human infusion trials, or have 
been epidemiological studies that only looked at total NEFA. To date, there have
been no longitudinal studies examining the role of the composition of the NEFA 
blood lipid fraction on the pathogenesis of diabetes, which is key to 
understanding causal associations given the chronic nature of diabetes 
pathophysiology. Therefore, our research objective was to examine the role of
the serum NEFA composition on insulin sensitivity and beta-cell function in a
longitudinal cohort. Some specific hypotheses included that higher palmitic acid
while lower polyunsaturated fatty acids such as eicosapentaenoic acid would 
associate with lower insulin sensitivity and beta-cell function.

@@

Participants from London and Toronto, Canada, were recruited into the 
Prospective Metabolism and Islet cell Evaluation (PROMISE) cohort. Eligibility 
for recruitment into PROMISE included having one or more risk factors for type 2
diabetes mellitus, such as obesity, hypertension, family history of diabetes, 
and/or a history of gestational diabetes or birth of a macrosomic infant. 
Participants aged 30 years and older (n=736) attended the baseline visit
between 2004-2006. Follow-up examinations occur every three years, with three
examination visits completed to date (2004-2006, 2007-2009, and 2010-2013).  The
current study used data on participants who did not have diabetes at baseline,
who returned for one or more follow-up examinations, and who had samples
available for fatty acid measurements (n=477). A diagram of the sample size at
each visit is shown in `r cite_sf('consort')`. At each examination,
participants undergo metabolic characterization, anthropometric measurements,
and questionnaires on lifestyle and sociodemographics. Research ethics approval
was obtained from Mount Sinai Hospital and the University of Western Ontario. 
Data collection methods were standardized across the 2 centres and research
nurses were centrally trained.

@@

NEFA composition was quantified using stored fasting serum samples from the 
baseline visit, which had been frozen at -70°C for 4-6 years and had not been 
exposed to any freeze-thaw cycles. Serum fatty acids have been documented to be
stable at these temperatures for up to 10 years [@Matthan2010a]. A known amount 
of heptadecanoic acid (17:0) {{confirm}} was added to the serum as an internal 
standard prior to extracting total lipids according to the method of Folch 
[@Folch1957a]. Each serum lipid fraction (NEFA, cholesteryl ester, phospholipid,
and triacylglycerol) was isolated using thin layer chromatography, which were
visualized under UV light after lightly spraying with 8-anilino-1-naphthalene
sulfonic acid (0.1% wt/vol) and then converted to fatty acid methyl esters with
14% boron trifluoride in methanol at 100°C for 1 h. Fatty acid methyl esters
were separated and quantified using a Varian-430 gas chromatograph (Varian, Lake
Forest, CA, USA) equipped with a Varian Factor Four capillary column and a flame
ionization detector, which were injected in splitless mode. Fatty acid
concentrations (nmol/ml) were calculated by proportional comparison of gas
chromatography peak areas to that of the internal standards [@Nishi2014a;
@Chen2011a].  There were 22 fatty acids measured in the NEFA fraction. Given the
diverse biology, as well as the complexity of the analyses, the other lipid
fractions were split into separate reports (see {{pl and ce paper}} for the
analysis of the phospholipid and cholesteryl ester fractions).

@@

For the primary analysis, generalized estimating equation (GEE) models 
[@Zeger1986a] were used to determine the longitudinal associations between the 
outcome variables and the predictor variables. Given the longitudinal design 
an auto-regressive (AR1) correlation matrix was chosen for the GEE models, 
though other matrices (eg. exchangeable) had similar fit (data not shown). GEE
is well suited to cohort studies given it's capacity to handle missed visits.
The predictor variables and continuous covariates were scaled (mean centered and
standardized). The NEFA variables were classified as *time-independent* (held
constant) as they were measured only at the baseline visit, while the outcome
variables and covariates were set as *time-dependent*. No imputation was
conducted on missing values.

Covariate selection was based on previous literature, directed acyclic graph 
[@Greenland1999a] recommendations, and quasi-likelihood information criterion 
(QIC). `r cite_st('qic')` shows the covariates compared using QIC and 
`r cite_sf('dag')` shows the specified directed acyclic graph. The final GEE model 
we selected differed between insulin sensitivity and beta-cell function models. 
For insulin sensitivity, we adjusted for time, sex, ethnicity, baseline age, WC,
ALT, MET, alcohol intake, and family history of diabetes (M6). For beta-cell
function, the last few models were comparable in performance, so we chose the
simplest model (M8), adjusting for time, sex, ethnicity, baseline age, WC, ALT,
and family history of diabetes. After scaling, log-transforming, and
exponentiating the GEE estimates, the GEE results are interpreted as an expected
percent difference in the outcome variable for every SD increase in the
predictor variable given the covariates are held constant. While models with a
time interaction did not perform better than others for the QIC (M6.1 and M8.1)
in `r cite_st('qic')`, we ran models with a time by predictor term for each of
the NEFA to test if there was an interaction by time. Given that TG is a risk factor
for diabetes and that NEFA contribute to TG production, TG may act as a mediator
between NEFA and the outcomes. However, only a subset of the participants at the
6-yr visit had clinically measured TG. To determine the role of TG in the
association between NEFA and the outcomes, TG was included in the GEE model as a
sensitivity analysis. Lastly for the GEE models, we tested for an interaction
with sex or ethnicity by the predictor term for the outcome variables.

While GEE controls for the longitudinal design of the data, a limitation is that
it can not analyze the inherent multivariate nature of the composition of the 
NEFA fraction. As such, after running the GEE models for the longitudinal 
analysis, partial least squares discriminant analysis (PLS-DA) was used to
identify the patterns of NEFA composition against beta-cell function (using ISSI-2,
which has most of the associations in the GEE models) and for those who converted or
maintained dysglycemia status (either IFG, IGT, or DM) over the 6 years. Since
PLS-DA can't handle longitudinal data, the longitudinal data was included by
extracting latent trajectories of beta-cell function (ISSI-2) over the 6 years using
latent class mixed models (LCMM). A limitation of these methods, excluding LCMM,
is that they can not be adjusted for covariates.

All analyses were performed using R `r getRversion()` [@Rbase], along with the R
packages geepack `r packageVersion('geepack')` for GEE [@Hoejsgaard2006a], caret
`r packageVersion('caret')` for PLS-DA, and lcmm `r packageVersion('lcmm')` for
LCMM. The R code for this manuscript is available at {{link}}. Results were
statistically significance at p<0.05, after adjusting for multiple testing using
the Benjamini-Hochberg False Discovery Rate [@Benjamini1995a].

@@

```{r variableAndDataSet, message=FALSE, results='hide'}
gee_df <- get_gee_data(ds)
dysgly_data <- get_dysglycemia_data(ds)

@@

```{r lcmmPls, cache=TRUE}
# include also the 3-yr and 6-yr data for training/testing.
lcmm_results <- analyze_lcmm(ds)
lcmm_data <- combine_lcmm_data(ds, lcmm_results)
plsda_results <- analyze_plsda(lcmm_data)
plsda_dysgly <- dysgly_data %>% 
    mutate(ConvertDysgly = factor(ConvertDysgly, levels = c(0, 1), labels = c('No', 'Yes'))) %>% 
    analyze_plsda('ConvertDysgly')
```

```{r inlineResults}
percent.nefa <- calculate_percent_nefa_contribution(ds) 
misclass.n.percent.bcf <- calculate_plsda_misclass(plsda_results)
misclass.n.percent.dysgly <- calculate_plsda_misclass(plsda_dysgly)
percent.change.outcomes <- calculate_outcomes_pct_change(ds)
lcmm.groups <- calculate_ngroups_lcmm(lcmm_results)
dm.convert <- calculate_conversion_dysgly(dysgly_data, 'ConvertDM')
predm.convert <- calculate_conversion_dysgly(dysgly_data, 'ConvertPreDM')
sex.npct <- calculate_factor_npercent(ds0yr$Sex, 'Female')
ethn.npct <- calculate_factor_npercent(ds0yr$Ethnicity, 'European')
fhd.npct <- calculate_factor_npercent(ds0yr$FamHistDiab, 'Yes')
est.gee <- calculate_gee_magnitude(gee_results)
```

@@

Within this subset of the PROMISE cohort at the baseline visit, 
`r sex.npct$npct` were females and `r ethn.npct$npct` had European-ancestry.
The average age of the participants was `r aide::average(ds0yr$Age)` years 
(`r aide::stddev(ds0yr$Age)` SD) and the average BMI was
`r aide::average(ds0yr$BMI)` (`r aide::stddev(ds0yr$BMI)` SD). Most of the 
subset of the cohort, `r fhd.npct$npct`, had a family history of diabetes.
The primary outcome variables had a significant median decline between 
`r percent.change.outcomes$chg` (p`r percent.change.outcomes$p`) over the 6
years in this subset of the PROMISE cohort.
Consistent with this decline, there were 42 (9%) <!--`r dm.convert`-->
participants who developed diabetes while `r predm.convert` participants either
converted to or maintained IFG or IGT status over the 6 years.

`r cite_f('nefaDist')` shows the distribution of the NEFA composition. Four 
individual NEFA made up the vast majority (`r sum(percent.nefa$pct)`%) of the total
NEFA fraction. The largest contributors were `r percent.nefa$c[1]`, 
`r percent.nefa$c[2]`, `r percent.nefa$c[3]`, and `r percent.nefa$c[4]`. All
individual NEFA as well as the total fraction had weak to no correlations
(r<0.3) with the basic characteristics (see `r cite_sf('corr')`).

@@

Several associations were seen in the unadjusted GEE models 
(see `r cite_sf('gee')`), particularly for the concentration data. Full model 
adjustment (`r cite_f('gee')`) attenuated most of these associations, where
total NEFA, 16:0, and 18:2n-6 (all as nmol/mL) had negative associations between
IGI/IR and ISSI-2. The magnitude of association for each of these variables was
fairly consistent for each beta-cell function measure. For every one SD decrease
in any of these three NEFA variables, there was an average predicted
`r est.gee$est[1]`% lower IGI/IR and `r est.gee$est[2]`% lower ISSI-2 at each
clinic visit. None of the NEFA variables as a mol% had any association with the
outcomes. Adjusting for TG attenuates all associations with IGI/IR and ISSI-2
(data not shown), though sample size at the 6 year visit is restricted to n=128
in those who had TG measured. There were no significant interaction effects
between time, sex, or ethnicity and the individual NEFA on any of the outcome
measures (data not shown). Raw values from the GEE models are shown in 
`r cite_st('gee_unadj')` for unadjusted models and `r cite_st('gee')` for fully
adjusted models.

@@

Three latent classes were extracted from LCMM, with `r lcmm.groups$Freq[2]` 
participants in the high, `r lcmm.groups$Freq[1]` in the middle, and 
`r lcmm.groups$Freq[3]`
in the low beta-cell function groups. The trajectories of the three groups are
shown in `r cite_sf('lcmm')`. These groups were used as the beta-cell function
response variables in the PLS-DA analysis. The PLS-DA results show a very poor
discriminatory ability of the NEFA composition in classifying participants into
the correct beta-cell function latent class. Only 
`r misclass.n.percent.bcf$NPct[2]` 
participants were correctly classified, primarily into the
middle group. The specific NEFA composition was not able to accurately classify
participants who had low or high beta-cell function.
`r cite_sf('plsda_groups')` shows the loadings of individual NEFA by the
extracted component and `r cite_sf('plsda_loadings')` shows the clustering of
the components by ISSI-2 latent class. The PLS-DA analysis using dysglycemia
status conversion or maintanence as the response variable showed similar poor 
discriminatory ability, though better than the beta-cell function trajectory
results. While `r misclass.n.percent.dysgly$NPct[2]` participants were
classified correctly based on dysglycemia status, examining the clustering of
the NEFA composition showed poor discrimination between groups (see
`r cite_sf('plsda_groups_dysgly')` and `r cite_sf('plsda_loadings_dysgly')`).

@@

In a Canadian population of adults who are at-risk for diabetes, we found that
higher total NEFA independently predicted a lower beta-cell function over 6 years.
While we found negative association with palmitic acid (16:0), oleic acid (18:1n-9), and
linoleic acid (18:2n-6) as a concentration (nmol/mL), no associations were seen
for the proportion of these fatty acids and further clustering analysis found no
predictive ability of these NEFA on beta-cell function or dysglycemia. These
observations suggest that the absolute size of the total fraction, rather than
the specific composition of any individual NEFA, likely influences the
pathogenesis of diabetes.

The role of total NEFA on diabetes is a well-studied area of research. 
Epidemiological studies have shown that higher NEFA associate with lower insulin
secretion and a higher risk for developing diabetes 
[@Salgin2012a;@Pankow2004a]. In a cross-sectional analysis of the RISCK cohort,
total NEFA had a negative association with insulin sensitivity and a
particularly strong negative association with beta-cell function [@Johns2014a].
Experimentally, several potential mechanisms have been elucidated for the role
of NEFA on beta-cell function, particularly for palmitic acid. Prolonged
exposure to elevated NEFA can induce apoptosis in the beta-cells, possibly
through endoplasmic reticulum stress, formation of ceramides, and generation of
nitric oxide, as well as impairment of proinsulin production and mitochrondial
function [@Cnop2008a; @Maris2013a; @Carpentier2003a; @Giacca2011a; @Xiao2009a].
The present analysis is the first study to our knowledge to examine a broad
spectrum of individual NEFA in a cohort setting on beta-cell function. We found
that there was a strong signal of higher total NEFA, palmitic acid, oleic acid,
and linoleic acid as concentrations with lower beta-cell function. However, as
both a mol% and using clustering analyses, no individual or specific
composition of the NEFA fraction predicted lower beta-cell function, insulin
sensitivity, or dysglycemia status. Together, these results suggest that the
absolute size of the NEFA fraction, irrespective of any specific composition of
fatty acid species, exerts it's hypothesized lipotoxic effects on the
beta-cells.
